---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `side::kick()`: A coding agent for RStudio

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/side)](https://CRAN.R-project.org/package=side)
[![R-CMD-check](https://github.com/simonpcouch/side/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/simonpcouch/side/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`side::kick()` is a coding agent for data science in RStudio, implemented entirely in R. It can interact with your files, talk to your active R session, and run code.

```{r}
#| echo: false
#| fig-alt: "A screencast of an RStudio session. A script tool-fetch_skill.R is open in the editor. A call to `side::kick()` in the console launches a non-blocking shiny chat app in a full-height viewer pane. The user then pastes a snippet of code into the chat dialog and says 'refactor this into a helper'. After calling multiple tools to locate the code, the chat app proposes two file edits in a git diff-based UI that refactor the code as desired."
knitr::include_graphics("https://github.com/user-attachments/assets/33e8d2bb-2e9e-4ff2-aef8-1695aaca3c77")
```


> [!Caution]
> 
> This project is highly experimental.
> 
> I put this app together to give myself a first pass at implementing several bits of functionality I'd like to see in ellmer/shinychat/btw. This package will not go to CRAN. Ideally, from here, more and more of the functionality prototyped in this app can make its way into ellmer/shinychat/btw, done "the right way," and ported back into this app. 


## Installation

To get started, install with:

```r
pak::pak("simonpcouch/side")
```

Then, run `side::kick()`. You might place `side::kick()` in your `.Rprofile` (perhaps with `usethis::edit_r_profile()`) to launch `side::kick()` every time you start R.

`sick::kick()` is intended for use with RStudio's *Sidebar*, a full-height pane that's currently only available in [RStudio Dailies](https://dailies.rstudio.com/). Install RStudio, then search for "Sidebar" in the command palette.

## Features

`side::kick()` can:

* **Use any LLM** supported by [ellmer](https://ellmer.tidyverse.org/)--Anthropic, OpenAI, Google Gemini, OpenRouter, and even local models are fair game.
* **See your active R session** to explore your computational environment.
* **Use tools** to read and write files, run shell commands, and read R documentation.
* **Create and execute plans**, allowing it to take on longer-horizon tasks.
* **Persist state** across R sessions so that you can close RStudio and come back to a chat later.
* Respond to **interrupts** so that you can stop execution and reroute the agent.
* Read **skills**, markdown files that describe how to carry out a given task in the way you need it to.

---

_`side::kick()` is implemented with [btw](https://posit-dev.github.io/btw/), [ellmer](https://ellmer.tidyverse.org/), and [shinychat](https://posit-dev.github.io/shinychat/)._
